Eco-evolutionary dynamics of finite populations from first principles

### 2023-02-26 ind_files/03_BD_processes.tex
Add ito formula in background

\section{Background and motivation}

Raymond Pearl, one of the pioneers of mathematical ecology as a discipline, wrote in 1927 that ``What we want to know is how the biological forces of natality and mortality are so integrated and correlated in their action as to lead to a final result in size of population which follows this particular curve rather than some other one'' \citep{pearl_growth_1976}. Pearl realized 95 years ago that population dynamics must be ultimately explained by the mechanistic processes of birth and death. Today, there are mounting calls for more mechanistic models of evolution that are grounded in these fundamental processes of birth and death \citep{geritz_mathematical_2012,doebeli_towards_2017}. Ecologists also increasingly recognize that incorporating stochasticity is vital to developing more realistic ecological models \citep{hastings_transients_2004, coulson_skeletons_2004, boettiger_noise_2018, shoemaker_integrating_2020,schreiber_does_2022} and does more than `add noise' to deterministic expectations. Individual-based models, where ecological rules are specified at the level of the individual, are a powerful mathematical tool for mechanistic descriptions of stochastic population dynamics \citep{black_stochastic_2012}. Birth-death processes are a very general class of stochastic processes that can be used to capture a wide range of eco-evolutionary processes. `System-size expansions' and their subsequent analysis using the `weak noise approximation' are common tools for analyzing stochastic birth-death processes that are well-known in the statistical physics and applied mathematics communities \citep{gardiner_stochastic_2009}. However, these tools remain relatively underappreciated in ecology, despite being relatively easy to understand and extremely well-motivated in scenarios germane to ecology and evolution. Here, I present a formulation of population dynamics constructed from first principles grounded in birth-death processes. To facilitate readership by a broad audience, only passing familiarity with calculus (derivatives, integrals, Taylor expansions) and probability are assumed. Familiarity with stochastic calculus is helpful for some sections but is not required, and I provide a brief introduction below. I begin by introducing birth-death processes, SDEs, and the Fokker-Planck equation in a non-technical manner, and providing the intuition for system-size expansions and weak noise analysis in ecological systems. In section \ref{sec_1D_processes}, I show how fluctuating population size of populations of identical individuals can be tracked through a one-dimensional birth-death process. I introduce a description of the system via a `master equation', and then conduct a `system-size expansion' to obtain a Fokker-Planck equation for the system. Finally, I conduct a weak noise approximation to arrive at a linear Fokker-Planck equation which can be solved exactly using some stochastic calculus to arrive at a closed-form solution given by a time-dependent Ornstein-Uhlenbeck process. As an example, I illustrate the complete process for a stochastic version of the logistic equation. In section \ref{sec_nD_processes}, I present a multivariate process to describe the evolution of discretely varying traits, and, as before, illustrate the system size expansion and the weak noise approximation. Under mild assumptions, we show that the deterministic limit of this process is the well-known replicator-mutator equation (or equivalently, Eigen's quasispecies equation), thus establishing the microscopic basis of well-known equations in evolutionary game theory. I also show that the mean value of the trait in the population changes according to the Price equation in the deterministic limit. Chapter \ref{chap_infD_processes} introduces a function-valued process to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via a system-size approximation to arrive at a `functional' Fokker-Planck equation, in terms of functional derivatives. Under mild assumptions, we show that classic equations such as Kimura's infinite alleles model and the canonical equation of adaptive dynamics can be derived as the deterministic limits of this stochastic process. We also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation.  Sections \ref{sec_1D_processes} and \ref{sec_nD_processes} apply techniques that are well-known in physics, as applied to population biology. Chapter \ref{chap_infD_processes} generalizes the work of Tim Rogers and colleagues \citep{rogers_demographic_2012,rogers_spontaneous_2012,rogers_modes_2015}, and to the best of my knowledge, is entirely original. Chapter \ref{chap_infD_processes} also presents a simple heuristic derivation of quantitative genetics models and adaptive dynamics from stochastic first principles that is much simpler than the rigorous mathematical derivations grounded in measure theory and martingale/markov theory that are currently standard reference in theoretical ecology \citep{champagnat_individual_2008}. I illustrate the utility of all this abstract formalism through examples in Chapter \ref{chap_examples}.

@@

where $\eta(t)$ is supposed to be `Gaussian white noise', defined indirectly such that $\int_0^{t}G(s,x)\eta(s)ds$ behaves the same as $\int_0^{t}G(s,X_s)dB_s$. However, it is important to remember that these are both purely formal expressions - Equation \eqref{ito_SDE_diff} is meaningless on its own and is really just shorthand for equation \eqref{ito_SDE_integral}, which is well-defined as explained above; Equation \eqref{ito_langevin} is even worse, because the Brownian motion is known to be non-differentiable, and as such, $\eta(t)$ cannot really exist - Both equations are thus to be interpreted as shorthand for equation \eqref{ito_SDE_integral}, which formally `makes sense'. SDEs are convenient because they satisfy several `nice' analytical properties. For example, using the fact that the Brownian motion has no expected change in value (\emph{i.e} $\mathbb{E}[B_t | B_0] = B_0 = 0$), it can be shown\footnote{We can actually prove something stronger: We can show under rather mild regularity assumptions on $X_t$ and $G(t,x)$ that the stochastic integral is a continuous square-integrable martingale starting at the origin - This means that the map $t \to \int_0^{t}G(s,X_s)dB_s$ is continuous, starts at the origin, and always has an expectation value of $0$.} that the stochastic integral also has an expectation value of $0$ for all $t$, \emph{i.e}:

@@

which becomes very useful for formal manipulation. Using this property, one can show using some simple algebra that if a process $X_t$ taking values in $\mathbb{R}$ satisfies the SDE \eqref{ito_SDE_diff}, then the probability density $P(x,t)$ of finding the process in a state $x \in \mathbb{R}$ at time $t$ satisfies the PDE

@@

. I present a simple informal derivation in Appendix \ref{App_SDE_FPE}. Equation \eqref{ito_FPE} is called the `Fokker-Planck equation' in the physics and applied mathematics literature \citep{gardiner_stochastic_2009} and is often called the `Kolmogorov forward equation' in the population genetics \citep{ewens_mathematical_2004} and pure mathematics \citep{oksendal_stochastic_1998} literature. If the function $G$ is independent of $x$, then it comes out of the derivatives in equation \eqref{ito_FPE}, and the resultant Fokker-Planck equation is said to be `linear' (and is much easier to solve). This link between SDEs and Fokker-Planck equations goes both ways: One can show that every stochastic process with a probability density described by a Fokker-Planck equation of the form \eqref{ito_FPE} corresponds to the solution of an SDE of the form \eqref{ito_SDE_diff}, though the proof is much more technical and will not be discussed here. This two-way correspondence proves to be extremely useful, as one approach often works for applications in which the other fails. This correspondence makes it greatly desirable to be able to describe our stochastic process of interest as either the solution to an It\^o SDE of the form \eqref{ito_SDE_diff} or as the solution to a Fokker-Planck equation of the form \eqref{ito_FPE}. System-size expansions facilitate such a description for birth-death processes.

@@

\section{Warm up: One-dimensional processes for population size}\label{sec_1D_processes}
The simplest birth-death processes are those in which the state at any time can be characterized by a single number. Populations of identical individuals are an obvious example of such a system. The mathematics below are adapted from sections 6.3 and 7.2 of \citep{gardiner_stochastic_2009}.

>>>>

\section{Mathematical Background}

Here, I provide a brief, informal introduction to basic notions in stochastic processes that will be helpful for some technical portions of this thesis. Interested readers looking for a more comprehensive source can refer to standard mathematics texts such as \cite{oksendal_stochastic_1998} or \cite{karatzas_brownian_1998} for a more rigorous treatment of the mathematical foundations, or physics-style texts such as \cite{gardiner_stochastic_2009} or \cite{van_kampen_stochastic_1981} for useful tools and techniques to study real systems.

@@

where $\eta(t)$ is supposed to be `Gaussian white noise', defined indirectly such that $\int_0^{t}G(s,x)\eta(s)ds$ behaves identically to $\int_0^{t}G(s,X_s)dB_s$. However, it is important to remember that these are both purely formal expressions - Equation \eqref{ito_SDE_diff} is meaningless on its own and is really just shorthand for equation \eqref{ito_SDE_integral}, which is well-defined as explained above; Equation \eqref{ito_langevin} is even worse, because the Brownian motion is known to be non-differentiable, and as such, $\eta(t)$ cannot really exist - Both equations are thus to be interpreted as shorthand for equation \eqref{ito_SDE_integral}, which formally `makes sense'. SDEs are convenient because they satisfy several `nice' analytical properties. For example, using the fact that the Brownian motion has no expected change in value (\emph{i.e} $\mathbb{E}[B_t | B_0] = B_0 = 0$), it can be shown\footnote{We can actually prove something stronger: We can show under rather mild regularity assumptions on $X_t$ and $G(t,x)$ that the stochastic integral is a continuous square-integrable martingale starting at the origin - This means that the map $t \to \int_0^{t}G(s,X_s)dB_s$ is continuous, starts at the origin, and always has an expectation value of $0$.} that the stochastic integral also has an expectation value of $0$ for all $t$, \emph{i.e}:

@@

which becomes very useful for formal manipulation. One important consequence is that we can no longer rely on the normal rules of calculus when dealing with stochastic integrals. In regular calculus, if we had a quantity $x(t)$ satisfying $\dot{x} = f(x,t) + g(x,t)$ for two `nice'\footnote{We just need the solution $x(t)$ to be a continuous function} real functions $f$ and $g$, then, given any function $h(x)$, we have the chain rule of differentiation, which says that
\frac{dh}{dt} = \frac{dh}{dx}\frac{dx}{dt} = h'(x)f(x) + h'(x)g(x)
or, in integral form,
	dh =  h'(x)f(x)dt + h'(x)g(x)dt
Naively, we may expect the same logic to still hold true for one-dimensional It\^o processes of the form \eqref{ito_SDE_diff} with $gdt$ simply being replaced by $GdB_t$ on the RHS. However, this intuition is not true. The correct relation is given by 
\emph{It\^o's formula}\footnote{It\^o's formula also additionally requires $h \in C^2(\mathbb{R})$, meaning that $h$ is continuous and the first and second derivatives of $h$ exist and are also continuous}, which states:
dh(X_t) = h'(X_t)F(X_t)dt + h'(X_t)G(X_t)dB_t  + \frac{h''(X_t)}{2}G^2(X_t)dt 
where the presence of the stochastic term has added an entire extra term $G^2(X_t)h''(X_t)/2$ that does not exist in the deterministic setting, in some sense because the Brownian motion `wiggles too erratically' over time. Using It\^o's formula and some simple algebra, one can then show that if a process $X_t$ taking values in $\mathbb{R}$ satisfies the SDE \eqref{ito_SDE_diff}, then the probability density $P(x,t)$ of finding the process in a state $x \in \mathbb{R}$ at time $t$ (if it exists) satisfies the PDE

@@

I present a simple informal derivation in Appendix \ref{App_SDE_FPE} for the sake of completeness. Equation \eqref{ito_FPE} is called the `Fokker-Planck equation' in the physics and applied mathematics literature \citep{gardiner_stochastic_2009} and is often called the `Kolmogorov forward equation' in the population genetics \citep{ewens_mathematical_2004} and pure mathematics \citep{oksendal_stochastic_1998} literature. If the function $G$ is independent of $x$, then it comes out of the derivatives in equation \eqref{ito_FPE}, and the resultant Fokker-Planck equation is said to be `linear' (and is much easier to solve). This link between SDEs and Fokker-Planck equations goes both ways: One can show that every stochastic process with a probability density described by a Fokker-Planck equation of the form \eqref{ito_FPE} corresponds to the solution of an SDE of the form \eqref{ito_SDE_diff}, though the proof is much more technical and will not be discussed here. This two-way correspondence proves to be extremely useful, as one approach often works for applications in which the other fails. This correspondence makes it greatly desirable to be able to describe our stochastic process of interest as either the solution to an It\^o SDE of the form \eqref{ito_SDE_diff} or as the solution to a Fokker-Planck equation of the form \eqref{ito_FPE}. System-size expansions facilitate such a description for birth-death processes.

@@

\section{Warm up: One-dimensional processes for population size}\label{sec_1D_processes}
The simplest birth-death processes are those in which the state at any time can be characterized by a single number. Populations of identical individuals are an obvious example of such a system. I will use this toy system as an illustration of the techniques that will be used for the actual problems we intend to tackle in the next sections. The mathematics below are adapted from sections 6.3 and 7.2 of \citep{gardiner_stochastic_2009}.
